Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-07-12 Delaneyetal2024.tex
small edits
changes on overleaf

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom.
  This quantity varies with water velocity and channel width.
  In subaerial channels,  sediment transport capacity can evolve consistently because water discharge variations are accommodated by changes in flow depth, width, and water velocity.
  By contrast, subglacial channels grow at a slower rate than most water discharge variations, causing a variable relationship between channel shape, water velocity, and discharge.
  Using numerical models applied to two observed hydrographs and sediment transport formulations, we evaluate the differences in sediment transport capacity variability between subaerial and subglacial channels.
  The numerical experiments confirm that subglacial sediment transport is highly non-linear with respect to water discharge.
  Yet, subglacial sediment transport capacity can vary less if water discharge variations are closer to equilibrium with subglacial channel size.
  In a dedicated numerical experiment, we show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  The results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  This implies that hysteresis in channel size causes a variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The implications are discussed for sediment discharge from glaciers with different hydro-climatic forcings.
  We also evaluate the conditions under which subglacial sediment transport capacity could co-vary with water discharge.
  These findings can also improve interpretation of sediment discharge records in glacierized catchments.
\end{abstract}

@@

Because of the above, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, several modeling frameworks examine subglacial sediment transport with evolving channel size \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
However, these works minimally discuss the impact of different hydrological regimes on sediment transport capacity underneath glaciers in the context of interpreting sediment transport records.
The few explicit parameterizations of subglacial sediment transport capacity with respect to water discharge assume fixed channel size  \citep{alley1997}. 
These formulations demonstrate a strongly non-linear response in subglacial sediment transport capacity to water discharge.
Yet, the effects of evolving channels size could impact the sediment transport response. 
Understanding these processes is imperative for establishing the effect of hydro-climatic conditions on subglacial sediment dynamics.
This is especially the case because sediment discharge capacity controls the mobilization and deposition of sediment, making it a fundamental aspect of subglacial sediment evacuation, especially over short timescales.

Despite the physical differences between the subaerial and subglacial systems, water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,hodson1998,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of these studies discuss the disparity between water discharge and sediment export, particularly the non-linear relationship between the two, along with the role of sediment access or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited, or hydraulically controlled, regime.
The response of sediment transport capacity to water discharge variations may affect the interpretation of sediment transport records in these glacierized catchment systems \citep[e.g.][]{ganti2016,mancini2023}.
This relationship must be clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

This manuscript has two objectives: 1) to evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs, and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
We use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
Lastly,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, steady-state R-channel, and pipe flow conditions.
These formulations extend the relationships presented in \citet{alley1997}, and illustrate the behavior of different channel types.
Findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel model accounts for the channel geometry and the water's velocity.
We use a lumped hydraulics model from \citet{werder2010b}, itself based upon  \citet{clarke1996}.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier.
    A second line refers to the range of values examined in the parameter search.}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
x    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

The first numerical experiment aims to compare the variability between the subglacial and subaerial model outputs to a range of channel shapes, friction factors, and water discharge variability.

Across the range of parameters examined (Section\,\ref{sect:imp}), variability in all model outputs remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, subglacial model outputs' variability is a magnitude larger than their subaerial counterparts.
Variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately $1$--$5$ days (Figure~\ref{fig:multi_run}).
These smoothing timescales remove the diurnal variations in water discharge, thereby reducing variability in model outputs. 
Velocity variability in \icesheet{} is substantially smaller than \alpine{} (Figure~\ref{fig:multi_run}\,a and \, d).
This could result from the subglacial conduit being in closer equilibrium with water discharge.
Variations in \icesheet{} shear stress are comparable to \alpine due to the effects the evolution of shear stress with subglacial conduit size (Equations ~\ref{eq:dS_dt} and \ref{eq:DW};). 
The much greater water flux and thus subglacial channel size in \icesheet{} drives the larger width integrated shear stress variations (Figure~\ref{fig:Qw}\,c and f).

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3.pdf}
  \caption{Standard deviation of detrended model outputs for different smoothing times with different subglacial and subaerial channel shapes and friction factors.
    Shaded areas denote the range of standard deviations from the accepted $100$ parameter combinations (Section~\ref{sect:imp}).
    Solid lines denote the mean value of standard deviations.
    Markers show smoothing periods ($15$\,\unit{min}, $1$\,\unit{hr}, $6$\,\unit{hr} $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d}).
  }
  \label{fig:multi_run}
\end{figure}

Smaller values of $\alpha$, or a channel cross-section closer to a slot canyon, result in greater variability in velocity and shear stress (Figure~\ref{fig:params}).
In both cases, the variability of small values of $\alpha$ does not approach that of the subglacial channels.
Yet, larger values of $\alpha$ can result in greater variability in width-integrated shear stress, as greater channel width changes increase the width-integrated shear stress.
Additionally, steeper subaerial channel slopes result in greater variability in both shear stress and width-integrated shear stress but only approach the subglacial channel's variability in the \alpine case (Figure~\ref{fig:params}). 
We note that these parameter values span a commonly accepted range.
Therefore, we do not anticipate a scenario where variability in the subaerial system would exceed the subglacial system with these two hydrologic forcings.

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig4.pdf}
  \caption{Parameter values compared to variability in model outputs.
    Plots on the right (orange, light blue) correspond to the \alpine{} case. Plots on the left (brown, dark blue) correspond to the \icesheet case.
    Outputs are shown using $15$ minute smoothing time. 
  }
  \label{fig:params}
\end{figure}

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=\linewidth]{Fig6.png}
  \caption{
    Relationship between water discharge and normalized velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet{} (d-f).
    Variables on the x-axis have been normalized to mean values.
    $R_g$ ($R_f$) shows the Spearman rank correlations for the subglacial (subaerial) outputs.
    Plots are shown with discharge and model outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

@@

\FloatBarrier
\section{Sensitivity of sediment transport in different drainage regimes}
\label{sect:scaling}

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.7\linewidth]{Fig7.pdf}
  \caption{Response of sediment transport capacity (orange line) and channel size (bottom row) to a water discharge event under different channel conditions.
    a) Pipe flow with fixed channel size described in Section~\ref{sect:scaling}.
    b) R-channel described in Section \ref{sect:sub_mode}.
    c) Subaerial channel described in Section~\ref{sect:fluv}.
    d) Steady state R-channel (channel size evolves in equilibrium with water discharge) described in Section~\ref{sect:scaling}.
    Note axes not to scale.
    Gyres on bottom panels represent shear stress on channel walls.
    $t_1$ ($t_2$) represents low (high) water discharge. 
  }
  \label{fig:chan_types}
\end{figure}

@@

\begin{table}[hbt!]
  \caption{Relations for hydraulic variables for the three drainage regimes:  subaerial channels, R-channel, and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \small
  \label{tab:eqs1}
  \begin{tabular}{llllll}
    Regime & Fixed & Determined via D-W
    & Additional relations & Shear stress & Stream p.\\
           & & (Equation~\ref{eq:DW})  &  &  \(\tau \propto\) & \(\Omega \propto\)\\
    \hline
    Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \makecell{\(w\,\propto Q^\alpha\) \\ \(S=wH\) \\ \(D_h\propto H\)} & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q\, \Psi\)\\
    R-channel & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q\, \Psi\)\\
    Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f\, Q^3 S^{-5/2}\)\\
  \end{tabular}
\end{table}

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the sediment transport capacity scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

@@

@@

While transport-limited states at glaciers are likely rare \citep[e.g.][]{alley1997}, abundant sediment could persist underneath some glaciers \citep[e.g.][]{walter2014,stevens2022,delaney2022}.
As a result, sediment export can be hydrologically controlled and efficient \citep[e.g.][]{swift2005,andresen2024}.
At these glaciers, the great variability in subglacial sediment transport capacity may make it difficult to link sediment discharge to hydrology, especially when peak events occur \citep{cowan1988,delaney2018,lu2022}.
The results here suggest that different channel sizes with the same hydrology forcing can result in very different sediment transport capacities, making it challenging to establish the effects of individual extreme precipitation or melt events.

R-channels rarely operate in a steady state with variations in water discharge, especially during severe rain or melt is too short for the channel to reach a steady state (Figures~\ref{fig:model_outs} and \ref{fig:Qw_vari}).
In these cases with high water discharge variability, channels can behave more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $\frac{7}{5}$.
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep{alley1997}, when evolving subglacial channel size is not considered (Figure~\ref{fig:model_outs}).
Note that pipe flow would cause sediment discharge capacity to covary with water discharge (Figure~\ref{fig:chan_types} a).
However, this covariance does not occur with evolving channel size (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types} b).

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge (Figure~\ref{fig:chan_types} b).
High water discharge variability in \alpine{} may cause shear stress to approach $Q^{2}$ in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Figure~\ref{fig:chan_types}; Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} and stronger correlation between width-integrated shear stress may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge (Figure~\ref{fig:chan_types} d).
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

The co-varying relationship between sediment discharge capacity and water discharge in subaerial conditions could be noticeable $\sim20$\,\unit{km} downstream of the Leverett site.
A strong correlation persists between sediment plume size and the Watson River's water discharge into the Kangerlussuaq fjord \citep[Figure~\ref{fig:chan_types}\,c][]{chu2009,mcgrath2010}.
In contrast, in marine-terminating glacier catchments, a less consistent relationship may occur between water discharge or melt extent and sediment plume size \citep{chu2012,tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \citep[e.g.][]{how2017}, so the observed reduced correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types}\,b).

@@

This study calls for the explicit consideration of rapidly evolving channel size when examining the relationship between sediment transport and hydro-climatic conditions from glacierized catchments, especially ones with high water discharge variability.

>>>>

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom.
  This quantity varies with water velocity and channel width.
  In subaerial channels,  sediment transport capacity can evolve consistently because water discharge variations are accommodated by changes in flow depth, width, and water velocity.
  By contrast, subglacial channels grow at a slower rate than most water discharge variations, causing a variable relationship between channel shape, water velocity, and discharge.
  Using numerical models applied to two observed hydrographs and sediment transport formulations, we evaluate the differences in sediment transport capacity variability between subaerial and subglacial channels.
  The numerical experiments confirm that subglacial sediment transport is highly non-linear with respect to water discharge.
  Yet, subglacial sediment transport capacity can vary less if water discharge variations are closer to equilibrium with subglacial channel size.
  In a dedicated numerical experiment, we show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  The results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  This implies that hysteresis in channel size causes a variable relationship between sediment and water discharge in a transport-limited subglacial system.
  Lastly, we present formulations that demonstrate the impact of different assumptions of channel behavior on sediment transport capacity.
  The implications are discussed for sediment discharge from glaciers with different hydro-climatic forcings.
  We also evaluate the conditions under which subglacial sediment transport capacity could co-vary with water discharge.
  These findings can also improve the interpretation of sediment discharge records in glacierized catchments.
\end{abstract}

@@

Because of the above, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, several modeling frameworks examine subglacial sediment transport with evolving channel size \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
However, these works minimally discuss the impact of different hydrological regimes on sediment transport capacity underneath glaciers in the context of interpreting sediment transport records.
The few explicit parameterizations of subglacial sediment transport capacity with respect to water discharge assume fixed channel size  \citep{alley1997}. 
These formulations demonstrate a strongly non-linear response in subglacial sediment transport capacity to water discharge.
Yet, the continually evolving channels' size can impact variations in sediment transport capacity.
Understanding these processes is imperative for establishing the effect of hydro-climatic conditions on subglacial sediment dynamics,
especially as sediment discharge capacity controls the mobilization and deposition of sediment.
This makes it a fundamental aspect of subglacial sediment evacuation, especially over short timescales.

Despite the different physical processes between the subaerial and subglacial systems, observed water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,hodson1998,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of these studies discuss the disparity between water discharge and sediment export, particularly the non-linear relationship between the two, along with the role of sediment access or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited, or hydraulically controlled, regime.
The response of sediment transport capacity to water discharge variations may affect the interpretation of sediment transport records in these glacierized catchments \citep[e.g.][]{ganti2016,mancini2023}.
This relationship must be clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019b} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

This manuscript has two objectives: 1) to evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs, and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
We use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
Lastly,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, steady-state R-channel, and pipe flow conditions.
These formulations extend the relationships presented in \citet{alley1997} and illustrate the sediment transport capacity behavior of different channel types.
Findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

The subglacial channel model accounts for the channel geometry and the water's velocity to evaluate the shear stress of water flowing across sediments underneath a glacier.
We use a lumped hydraulics model from \citet{werder2010b}, itself based upon  \citet{clarke1996}.

@@

\begin{table}[hbt!]
  \centering
  \caption{Variables, parameters, and constants used in this work.
    Where two values are given, the first refers to  \alpine{}, a scenario from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a scenario from Leverett Glacier.
    A second line refers to the range of values examined in the parameter search.}
  \small 
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\
         && (\alpine{} or \icesheet{})\\
    \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (subglacial, subaerial)  & $v$, ($v_g,\,v_{f}$)& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (subglacial, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Channel depth (subaerial) & $H$&& $\mathrm{m}$\\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (subglacial, subaerial) & $w$, ($w_g,w_f$)&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Hydraulic gradient &$\Psi=\frac{\Delta h}{l}$&& $\mathrm{m\, m^{-1}}$\\
    Shear stress (subglacial, subaerial) & $\tau$, ($\tau_g,\,\tau_f$) && $\mathrm{Pa \, m^{-2}}$ \\
    Stream power & $\Omega$ && $\mathrm{ kg \, m\, s^{-3}}$ \\

@@

The first numerical experiment aims to compare the variability between the subglacial and subaerial model outputs to a range of channel shapes, friction factors, and water discharge variability. 
This is accomplished by applying a range of parameter values to models applied to the different hydrological regimes and examining the variability in model outputs (Section~\ref{sect:imp}).

Across the range of parameters examined, variability in all model outputs (i.e. velocity, shear stress, and width-integrated shear stress) remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, subglacial model outputs' variability is a magnitude larger than their subaerial counterparts.
Variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately $1$--$5$ days (Figure~\ref{fig:multi_run}).
These smoothing timescales remove the diurnal variations in water discharge, thereby reducing variability in model outputs. 
Velocity variability in \icesheet{} is substantially smaller than \alpine{} (Figure~\ref{fig:multi_run}\,a and \, d).
This could result from the subglacial conduit being in closer equilibrium with water discharge.
Variations in \icesheet{} shear stress are comparable to \alpine due to the effects the evolution of shear stress with subglacial conduit size (Equations ~\ref{eq:dS_dt} and \ref{eq:DW};). 
The much greater water flux and thus subglacial channel size in \icesheet{} drives the larger width integrated shear stress variations (Figures~\ref{fig:Qw}~and~\ref{fig:Qw}\,c and f).

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.7\linewidth]{Fig3.pdf}
  \caption{Standard deviation of detrended model outputs for different smoothing times with different subglacial and subaerial channel shapes and friction factors.
    Shaded areas denote the range of standard deviations from the accepted $100$ parameter combinations (Section~\ref{sect:imp}).
    Solid lines denote the mean value of standard deviations.
    Markers show smoothing periods ($15$\,\unit{min}, $1$\,\unit{hr}, $6$\,\unit{hr} $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d}).
  }
  \label{fig:multi_run}
\end{figure}

Smaller values of $\alpha$, or a channel cross-section closer to a slot canyon, result in greater variability in velocity and shear stress (Figure~\ref{fig:params}).
In both cases, the variability of small values of $\alpha$ does not approach that of the subglacial channels.
Yet, larger values of $\alpha$ can result in greater variability in width-integrated shear stress, as greater channel width changes increase the width-integrated shear stress.
Additionally, steeper subaerial channel slopes $\nabla z_c$ result in greater variability in both shear stress and width-integrated shear stress but only approach the subglacial channel's variability in the \alpine case (Figure~\ref{fig:params}). 
We note that these parameter values span a commonly accepted range.
Therefore, we do not anticipate a scenario where variability in the subaerial system would exceed the subglacial system with these two hydrologic forcings.

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.7\linewidth]{Fig4.pdf}
  \caption{Parameter values compared to variability in model outputs.
    Plots on the right (orange, light blue) correspond to the \alpine{} case. Plots on the left (brown, dark blue) correspond to the \icesheet case.
    Outputs are shown using $15$ minute smoothing time. 
  }
  \label{fig:params}
\end{figure}

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.8\linewidth]{Fig6.png}
  \caption{
    Relationship between water discharge and normalized velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet{} (d-f).
    Variables on the x-axis have been normalized to mean values.
    $R_g$ ($R_f$) shows the Spearman rank correlations for the subglacial (subaerial) outputs.
    Plots are shown with discharge and model outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

@@

\FloatBarrier
\subsection{Sensitivity of sediment transport in different channel types}
\label{sect:scaling}

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.7\linewidth]{Fig7.pdf}
  \caption{Response of sediment transport capacity (orange line) and channel size (bottom row) to a water discharge event (blue line) under different channel conditions.
    a) Pipe flow with fixed channel size described in Section~\ref{sect:scaling}.
    b) R-channel described in Section \ref{sect:sub_mode}.
    c) Subaerial channel described in Section~\ref{sect:fluv}.
    d) Steady-state R-channel (channel size evolves in equilibrium with water discharge) described in Section~\ref{sect:scaling}.
    Note axes not to scale.
    Gyres on bottom panels represent shear stress on channel walls.
    $t_1$ ($t_2$) represents low (high) water discharge. 
  }
  \label{fig:chan_types}
\end{figure}

@@

\begin{table}[hbt!]
  \caption{Relations for hydraulic variables for the three different channel types:  subaerial channels, R-channel, and pipe-flow.  Darcy-Weisbach equation is abbreviated with ``D-W'', and stream power with ``Stream p.''. }
  \small
  \label{tab:eqs1}
  \begin{tabular}{llllll}
    Channel type & Fixed & Determined via D-W
    & Additional relations & Shear stress & Stream p.\\
           & & (Equation~\ref{eq:DW})  &  &  \(\tau \propto\) & \(\Omega \propto\)\\
    \hline
    Subaerial & \(\Psi\) & \(H \propto f^{1/3}\, Q^{2/3-2\alpha/3} \, \Psi^{-1/3}\) & \makecell{\(w\,\propto Q^\alpha\) \\ \(S=wH\) \\ \(D_h\propto H\)} & \(f^{1/3} Q^{2/3-2\alpha/3}  \Psi^{2/3}\) & \(Q\, \Psi\)\\
    R-channel & \(\Psi\) & \(S\, \propto f^{2/5}\, Q^{4/5} \, \Psi^{-2/5}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f^{1/5} Q^{2/5} \, \Psi^{4/5}\) & \(Q\, \Psi\)\\
    Pipe & \(S\) & \(\Psi \propto f \, Q^2\, S^{-5/2}\) & \(D_h\propto w \propto H \propto S^{1/2}\) & \(f Q^2 S^{-2}\) & \(f\, Q^3 S^{-5/2}\)\\
  \end{tabular}
\end{table}

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different channel types and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow channel type.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the sediment transport capacity scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

@@

R-channels rarely operate in a steady state with variations in water discharge, especially during severe rain or melt is too short for the channel to reach a steady state (Figures~\ref{fig:model_outs} and \ref{fig:Qw_vari}).
In these cases with high water discharge variability, channels can behave more like a pipe of fixed cross-section \citep[\alpine{} case in Figure \ref{fig:Qw_vari}; e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two channel types when that exponent is at most $\frac{7}{5}$.
Thus fluctuations of discharge on short timescales (on the order of a day; Figure~\ref{fig:multi_run}) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep[][]{alley1997}.
Note that pipe flow would cause sediment discharge capacity to covary with water discharge (Figure~\ref{fig:chan_types} a).
However, this covariance does not occur with evolving channel size (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types} b).

@@

While transport-limited states at glaciers are likely rare \citep[e.g.][]{alley1997}, abundant sediment could persist underneath some glaciers, potentially creating a transport-limited regime \citep[e.g.][]{walter2014,stevens2022,delaney2022}.
At these glaciers, the great variability in subglacial sediment transport capacity may make it difficult to link sediment discharge to hydrology, especially when peak events occur \citep{cowan1988,delaney2018,lu2022}.
The results here suggest that different channel sizes with the same hydrology forcing can result in very different sediment transport capacities, making it challenging to establish the effects of individual extreme precipitation or melt events.

Catchments with reduced water discharge variability may experience yet less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and sporadic relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge (Figure~\ref{fig:chan_types} b).
High water discharge variability in \alpine{} may cause width integrated shear stress to approach $Q^{2}$ in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Figure~\ref{fig:chan_types}; Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} and stronger correlation between width-integrated shear stress may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge (Figure~\ref{fig:chan_types} d).
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

The co-varying relationship between sediment discharge capacity and water discharge in subaerial conditions could be noticeable $\sim20$\,\unit{km} downstream of the Leverett site.
A strong correlation persists between sediment plume size and the Watson River's water discharge into the Kangerlussuaq fjord \citep[Figure~\ref{fig:chan_types}\,c; ][]{chu2009,mcgrath2010}.
In contrast, in marine-terminating glacier catchments, a less consistent relationship may occur between water discharge or melt extent and sediment plume size \citep{chu2012,tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \citep[e.g.][]{how2017}, so the observed reduced correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types}\,b).

@@

This study calls for the explicit consideration of evolving channel size when examining the relationship between sediment transport and hydro-climatic conditions from glacierized catchments, especially ones with high water discharge variability.
